Formalization of Forcing in Isabelle/ZF

### 2020-06-17 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
full account of ZF-Constructible, more details on achievements

  It makes sense to compare, for $x,y\in M$:
  \[
  x \sbq y
  \begin{overlayarea}{4.5em}{2ex}
    \ \ \ \only<3-4>{\ \  and}\only<5>{\alert{$\iff$}}
  \end{overlayarea}
  M \models x\sbq y
  \]
  \pause
  The right hand side can be written as a term of type \isatt{o}:
  $\forall z.\ z\in M \longrightarrow (z\in x \longrightarrow z\in
  y)$,
  the \textbf{relativization} of $\sbq$ to $M$. 
  \pause

  In this case, we know that  $\sbq$  is \textbf{absolute}.
  %
  %% \begin{block}{Forcing notions and generics}
  %%   \begin{itemize}
  %%   \item $\lb \PP, {\preceq} ,\1\rb$ preorder with top.
  %%   \item A filter $G\sbq \PP$ is \textbf{$M$-generic} if it meets
  %%     every dense $D\sbq \PP$ in $M$.
  %%   \end{itemize}
  %% \end{block}
\end{frame}

@@

  Every $a\in M[G]$ is coded by
  some $\punto{a}\in M$ through the $M$-definable function $\val$:
  \[
  M[G]\defi \{ \val(G,\punto{a}) : \punto{a}\in M\}
  \]
  \emph{Fundamentally}, \textbf{truth} in $M[G]$ is coded in $M$ by the
  function $\forceisa$.
  \pause
  \begin{theorem}[\citep{Cohen-CH-PNAS}]
    \begin{enumerate}
    \item $\bigl(\forall G\visible<4->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
      \models \phi\bigr)
      \iff
      M,
      [\only<-3>{\1}\alert{\only<4->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
      \forceisa(\phi)$.
    \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
      \iff M,$}\visible<4->{%
      $\overset{\searrow}{}$ \ 
      $\alert{p} \forces^M_{\PP,\preceq} \phi(\punto{a})$
      \  $\overset{\swarrow}{}$}
      \pause% 
      \pause% 
    \item For $env\in\mathsf{list}(M)$,
      \[
      M[G], \mathrm{map}(\val(G),env) \models \phi
      \iff
      \exists p\in G.\ M, [p,\preceq,\PP] \mathbin{{+}\!\!\!\!{+}} env \models
      \forceisa(\phi).
      \]
    \end{enumerate}
  \end{theorem} 
\end{frame}

\begin{frame}
  \frametitle{The ctm approach to forcing, 3/4}
  \begin{block}{Benefits of using ctms}
    \begin{enumerate}
    \item<+-> Countability ensures that generics exist (by
      Rasiowa-Sikorski Lemma).
      %  
    \item<+-> Absoluteness provide seamless
      treatment of many concepts.
      \begin{center}
        $\alpha$ is an ordinal \ $\iff$ \  $M\models \alpha$ is
        an ordinal \  $\iff$ \  $M[G]\models \alpha$ is an ordinal
      \end{center}
      %
    \item<+-> Both $M$ and $M[G]$ are standard (two-valued) models.
      %
    \item<+-> Ctms are
      used in an important (perhaps the greatest) fraction of the literature.
    \end{enumerate}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The ctm approach to forcing, 4/4}
  By choosing  $\lb \PP, {\preceq} ,\1\rb$
  appropriately one can tune the first order properties of  $M[G]$ (for
  any generic $G$).
  \pause
  %
  \begin{theorem}[\citep{Cohen-CH-PNAS}]
    If $\PP$ is the set of finite partial binary functions with domain
    included in $\aleph_2$, $M[G]$ satisfies the negation of the
    \textbf{Continuum Hypothesis} ($\CH$):
    \[
    M[G] \models 2^{\aleph_0} >\aleph_1.
    \]
  \end{theorem}
  \pause

@@

\begin{frame}
  \frametitle{What did we accomplish?}
  %
  \begin{shadowblock}{}
    \begin{enumerate}
    \item<1-> We adapted \texttt{ZF-Constructible} to obtain
      absoluteness results for nonempty transitive classes ($\leftarrow$
      \alert{in Isabelle2020}).

      \visible<2->{
        Around 40 absoluteness/closure lemmas now hold using weaker
        hypotheses on the class $C$ (most of them, just that $C$ is
        transitive and nonempty).
      }
    \item<3-> We formalized the formula transformer $\forceisa$ and
      hence the forcing relation $\forces$, and proved the Fundamental
      Theorems.
    \item<4-> We showed that generic extensions of ctms of $\ZF$ are also
      ctms of $\ZF$ (respectively, adding $\AC$).
    \item<5-> We provided the forcing notion that adds a Cohen real,
      therefore proving the existence of a nontrivial extension.
    \end{enumerate}
  \end{shadowblock}
\end{frame}

\begin{frame}
  \frametitle{Some details}
  \begin{enumerate}\setcounter{enumi}{1}
  \item<+-> We formalized the formula transformer $\forceisa$ and
    hence the forcing relation $\forces$, and proved the Fundamental
    Theorems.
    \begin{itemize}
    \item<+-> $\forceisa$ is defined by induction on internalized
      formulas.
    \item<+-> The base cases  $\forceisa(x\in y)$ and  $\forceisa(x =
      y)$ are  defined by wellfounded
      recursion, which \alert{must also be internalized}.
    \end{itemize}
  \item<+-> We showed that generic extensions of ctms of $\ZF$ are also
    ctms of $\ZF$ (respectively, adding $\AC$).
    \begin{itemize}
    \item<+-> We may compare some of the code with the actual
      math \citep{kunen2011set}.
    \end{itemize}
  \end{enumerate}
\end{frame}

>>>>

  It makes sense to compare, for $x,y\in M$:
  \[
  x \sbq y
  \begin{overlayarea}{4.5em}{2ex}
    \ \ \ \only<3-4>{\ \  and}\only<5>{\alert{$\iff$}}
  \end{overlayarea}
  M \models x\sbq y
  \]
  \pause
  The rhs can be written as a term of type \isatt{o}:

  {} \hfill   $\forall z.\ z\in M \longrightarrow (z\in x \longrightarrow z\in
  y)$,
  \hfill the \textbf{relativization} $\sbq^M$ of $\sbq$ to $M$. 
  \pause

  In this case, we know that  $\sbq$  is \textbf{absolute for
    transitive models}.
  %
  %% \begin{block}{Forcing notions and generics}
  %%   \begin{itemize}
  %%   \item $\lb \PP, {\preceq} ,\1\rb$ preorder with top.
  %%   \item A filter $G\sbq \PP$ is \textbf{$M$-generic} if it meets
  %%     every dense $D\sbq \PP$ in $M$.
  %%   \end{itemize}
  %% \end{block}
\end{frame}

@@

  Every $a\in M[G]$ is coded by
  some $\punto{a}\in M$ through the $M$-definable function $\val$:
  \[
  M[G]\defi \{ \val(G,\punto{a}) : \punto{a}\in M\}
  \]
  Fundamentally, \textbf{truth} in $M[G]$ is coded in $M$ by the
  function $\forceisa$.
  \pause
  \begin{theorem}[\citep{Cohen-CH-PNAS}]
    \begin{enumerate}
    \item $\bigl(\forall G\visible<4->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
      \models \phi\bigr)
      \iff
      M,
      [\only<-3>{\1}\alert{\only<4->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
      \forceisa(\phi)$.
    \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
      \iff M,$}\visible<4->{%
      $\overset{\searrow}{}$ \ 
      $\alert{p} \forces^M_{\PP,\preceq} \phi(\punto{a})$
      \  $\overset{\swarrow}{}$}
      \pause% 
      \pause% 
    \item For $env\in\mathsf{list}(M)$,
      \[
      M[G], \mathrm{map}(\val(G),env) \models \phi
      \iff
      \exists p\in G.\ M, [p,\preceq,\PP] \mathbin{{+}\!\!\!\!{+}} env \models
      \forceisa(\phi).
      \]
    \end{enumerate}
  \end{theorem} 
\end{frame}

\begin{frame}
  \frametitle{The ctm approach to forcing, 3/4}
  \begin{block}{Benefits of using ctms}
    \begin{enumerate}
    \item<+-> Countability ensures that generics exist (by
      Rasiowa-Sikorski Lemma).
      %  
    \item<+-> Absoluteness provides seamless
      treatment of many concepts.
      \begin{center}
        $\alpha$ is an ordinal \ $\iff$ \  $M\models \alpha$ is
        an ordinal \  $\iff$ \  $M[G]\models \alpha$ is an ordinal
      \end{center}
      %
    \item<+-> Both $M$ and $M[G]$ are standard (two-valued) models.
      %
    \item<+-> Ctms are
      used in an important (perhaps the greatest) fraction of the literature.
    \end{enumerate}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The ctm approach to forcing, 4/4}
  By choosing  $\lb \PP, {\preceq} ,\1\rb$
  appropriately one can tune the first order properties of  $M[G]$ (for
  any generic $G$).
  \pause
  %
  \begin{theorem}[\citep{Cohen-CH-PNAS}]
    If $\PP$ is the set of finite partial binary functions with domain
    included in $\aleph_2^M$, $M[G]$ satisfies the negation of the
    \textbf{Continuum Hypothesis} ($\CH$):
    \[
    M[G] \models 2^{\aleph_0} >\aleph_1.
    \]
  \end{theorem}
  \pause

@@

\begin{frame}
  \frametitle{What did we accomplish?}
  %
  \begin{shadowblock}{}
    \begin{enumerate}
    \item<1-> We adapted the \texttt{ZF-Constructible} library
      \citep{paulson_2003} to obtain
      absoluteness results for nonempty transitive classes ($\to$
      \alert{Isabelle2020}).
    \item<2-> We formalized the formula transformer $\forceisa$ and
      hence the forcing relation $\forces$, and proved the Fundamental
      Theorems.
    \item<3-> We showed that generic extensions of ctms of $\ZF$ are also
      ctms of $\ZF$ (respectively, adding $\AC$).
    \item<4-> We provided the forcing notion that adds a Cohen real,
      therefore proving the existence of a nontrivial extension.
    \end{enumerate}
  \end{shadowblock}
\end{frame}

\begin{frame}
  \frametitle{Some details}
  \begin{enumerate}%\setcounter{enumi}{1}
  \item We adapted \texttt{ZF-Constructible} to obtain
    sharper absoluteness results.
  \end{enumerate}

  This library contains a major development of relativization and
  absoluteness for \emph{classes} $C :: {}$ \isatt{i} $\Rightarrow$
  \isatt{o}, including the definition of the set \isatt{formula} ::
  \isatt{i} of internal formulas and model theoretic satisfaction. 
  \pause
  \begin{block}{Relativization and synthesis discipline}
    \begin{enumerate}[I]
    \item<+-> $p=\{x,y\} :: {}$\isatt{i}
      \hfill
      (original term).
    \item<+-> \isatt{upair}$(C,x,y,p) :: {}$\isatt{o}
      \hfill
      (relativization, fully
      relational). 
    \item<+-> \isatt{upair{\uscore}fm}$(0,1,2) :: {}$\isatt{i} 
      \hfill
      (synthesized member of \isatt{formula}).
    \end{enumerate}
  \end{block}
  \begin{alertblock}{}<+->
    Around 40 absoluteness/closure lemmas now hold using weaker
    hypotheses on the class $C$ (most of them, just that $C$ is
    transitive and nonempty).
  \end{alertblock}
\end{frame}

\begin{frame}{Some details}
  \begin{enumerate}\setcounter{enumi}{1}
  \item<+-> We formalized the formula transformer $\forceisa$ and
    hence the forcing relation $\forces$, and proved the Fundamental
    Theorems.
  \end{enumerate}
  \begin{itemize}[I]
  \item<+-> $\forceisa$ is defined by induction on internalized
    formulas.
  \item<+-> The base cases  $\forceisa(x\in y)$ and  $\forceisa(x =
    y)$ are  defined by wellfounded
    recursion, which \alert{must also be internalized}.
  \end{itemize}
\end{frame}

\begin{frame}{Some details}
  \begin{enumerate}\setcounter{enumi}{2}
  \item For any ctm $M$ of $\ZF$ and $M$-generic $G$, $M[G] \models
    \ZF$ (respectively, adding $\AC$).
  \end{enumerate}
  \pause
  Actually, the modularity of the theory of forcing allowed us to do
  this \alert{before} we write down the definition of $\forceisa$.
  \pause

  We may compare some of the code with the actual
  math \citep{kunen2011set}.
\end{frame}
